Individual and Differential Harm in Redistricting

### 2025-04-28 paper/harm.qmd
fix app refs too

Motivated primarily by computational convenience and simplicity, we adopt the stochastic uniform partisan swing model of @gelman1994unified, modeling the Democratic two-party vote share in district $d$ and election $v$ as $$
    s_{dv} = \bar{s}_d + \delta_v + \varepsilon_{dv},
$$ where $\bar{s}_d$ is the baseline Democratic vote share in the districts, $\delta_v\sim\Norm(0, \sigma^2_\delta)$ is a election-specific uniform shift, and $\varepsilon_{dv}\sim\Norm(0, \sigma^2_\varepsilon)$ is a district-election residual error term.
The baseline $\bar{s}_d$ can be established by averaging the results of all available statewide elections over the past decade or other suitable period.
Since $\bar{s}_d$ is calculated for the particular configuration of district $d$, this model takes into account the specific districting plan under consideration.
We estimate $\sigma^2_\delta$ and $\sigma^2_\varepsilon$ by fitting a linear mixed-effects model to 5,931 contested U.S. congressional elections from 1976 to 2020.
Appendix @app:elecmodel contains more details on how we estimate and apply this model.

@@

Figure @fig-nj-pairs in the appendix expands on Figure @fig-nj-meas to also show the correlations between these various measures.
Expected Democratic seats, the differential harm estimates, the efficiency gap, and declination are all moderately correlated with each other.
Similarly, partisan bias and the mean-median score are moderately correlated, but very weakly correlated with the expected number of Democratic seats.
Thus there exist many plans which give Democrats a higher-than-average share of the seats, but which have partisan bias and mean-median score favoring Republicans.
New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
Appendix @app:corr records these correlations across simulated plans in eight states.
The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which averages 0.79 in the eight states we study.
Seats are, of course, more discrete and cannot directly identify the local areas where gains and losses are derived.

@@

As discussed in Section @sec-racial, plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect district could be drawn while meeting all other redistricting requirements.
In *Merrill,* the plaintiff's expert, Moon Duchin, submitted four such remedial plans, one of which is shown in Figure @fig-al-maps(e).
These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure @fig-al-maps(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
In Appendix @app:alsampling, we repeat the analysis performed here on a different counterfactual consisting of plans sampled with a redistricting simulation algorithm; the results are broadly similar.

@@

This analysis is slightly more complicated than in New Jersey, since in order to understand the racial distribution of harm we need to estimate voting preferences by race.
As is custom in VRA analyses, we use an ecological inference model.
The model, detailed in Appendix @app:alei, jointly estimates two turnout and vote choice by race and precinct across seven statewide elections using a setup following @king1997solution.
For example, we model precinct $j$'s observed turnout $n_{jv}$ in election $v$ and baseline turnout rates $\turn_{jr}$ by race $r$ as \begin{align*}
    n_{jv} \mid \turn_{jr}, \beta^{(t)}_v
        &\sim \Binlog\qty(
        \overbrace{\vap_j,}^{\substack{\text{voting-age}\\\text{population}}}
        \hspace{0.5em}
        \logit(\turn_j^\top
        \overbrace{\vb p_j}^{\mathclap{\substack{\text{precinct racial}\\\text{proportions}}}}) + 
        \underbrace{\beta^{(t)}_v}_{\mathllap{\text{election-specific turnout shift}}}) \\
    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
        &\sim \mathcal{N}\underbrace{\qty(\mu^{(t)}, \Sigma^{(t)})}_{\mathrlap{
        \text{where to cluster the precinct turnout rates}}},
\end{align*} where $\Binlog$ indicates the Binomial distribution with the probability parameter specified on the logit scale.
We model Democratic support rates $\supp_{jr}$ by precinct and race similarly.

@@

Figure @fig-boston-factors shows the key outputs from this model, and Appendix @app:lfei explains the model in more detail.
As expected one dimension (the first) primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders, Elizabeth Warren, and current Boston mayor Michelle Wu on the left, and centrists like Joseph R. Biden and Wu's opponent, Annissa Essaibi-George, on the right.
The second dimension appears to capture racial differences in support---as the factor value maps in Figure @fig-boston-factors show, minority-majority and White areas of the city take opposite values on the second dimension.

Like all ecological inference models, ours relies on several unverifiable assumptions.
The latent factor component imposes additional assumptions about the structure of voting in Boston elections.
As Appendix @app:lfei discusses, these assumptions are almost certainly not *strictly* met.
However, the model is a useful simplification of reality that captures the major electoral dynamics relevant to our question of interest.
For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure @fig-boston-gen-pred in Appendix @app:lfei).
The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
We stress that the latent factor ecological inference model is not necessary for this kind of analysis; it is one out of several approaches which could provide estimated vote choices for individuals.
Local survey data with ideological preferences and racial information would be a sufficient alternative.

@@

```{=tex}
\vspace{1in}
\appendix
\renewcommand\thefigure{\thesection.\arabic{figure}}
\renewcommand\thetable{\thesection.\arabic{table}}
\setcounter{figure}{0}
\setcounter{table}{0}
```
# Two-party Election Model {#app:elecmodel}

@@

# Correlation Between Partisan Gerrymandering Metrics Across Multiple States {#app:corr}

@@

```{=tex}
\begin{table}[ht!]
\centering
\input{figures/corr_table.tex}
\vspace{0.1in}
\caption{Correlations between five gerrymandering metrics and partisan differential harm, for simulated districting plans from eight states. Declination is NA for Massachusetts since there are no Republican-leaning districts at baseline under any simulated plan.}
\label{tab:corr}
\end{table}
```
# Ecological Inference Model for Alabama {#app:alei}

@@

# Additional Sampling Analysis for Alabama {#app:alsampling}

@@

# Latent Factor Ecological Inference Model for Boston Primary Elections {#app:lfei}

To estimate voter preferences by race and precinct in Boston, we use 2020 U.S.
Census demographic data and election returns from the March 2020 presidential primary and the September 2021 mayoral preliminary elections.
Our ecological inference model used to produce the estimates expands on the model in Appendix @app:alei, jointly estimating both turnout and vote choice in the two elections.
However, it estimates voter preferences in a two-dimensional latent space and also estimates candidates' positions in this space.

>>>>

Motivated primarily by computational convenience and simplicity, we adopt the stochastic uniform partisan swing model of @gelman1994unified, modeling the Democratic two-party vote share in district $d$ and election $v$ as $$
    s_{dv} = \bar{s}_d + \delta_v + \varepsilon_{dv},
$$ where $\bar{s}_d$ is the baseline Democratic vote share in the districts, $\delta_v\sim\Norm(0, \sigma^2_\delta)$ is a election-specific uniform shift, and $\varepsilon_{dv}\sim\Norm(0, \sigma^2_\varepsilon)$ is a district-election residual error term.
The baseline $\bar{s}_d$ can be established by averaging the results of all available statewide elections over the past decade or other suitable period.
Since $\bar{s}_d$ is calculated for the particular configuration of district $d$, this model takes into account the specific districting plan under consideration.
We estimate $\sigma^2_\delta$ and $\sigma^2_\varepsilon$ by fitting a linear mixed-effects model to 5,931 contested U.S. congressional elections from 1976 to 2020.
Appendix @sec-app-elecmodel contains more details on how we estimate and apply this model.

@@

Figure @fig-nj-pairs in the appendix expands on Figure @fig-nj-meas to also show the correlations between these various measures.
Expected Democratic seats, the differential harm estimates, the efficiency gap, and declination are all moderately correlated with each other.
Similarly, partisan bias and the mean-median score are moderately correlated, but very weakly correlated with the expected number of Democratic seats.
Thus there exist many plans which give Democrats a higher-than-average share of the seats, but which have partisan bias and mean-median score favoring Republicans.
New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
Appendix @sec-app-corr records these correlations across simulated plans in eight states.
The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which averages 0.79 in the eight states we study.
Seats are, of course, more discrete and cannot directly identify the local areas where gains and losses are derived.

@@

As discussed in Section @sec-racial, plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect district could be drawn while meeting all other redistricting requirements.
In *Merrill,* the plaintiff's expert, Moon Duchin, submitted four such remedial plans, one of which is shown in Figure @fig-al-maps(e).
These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure @fig-al-maps(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
In Appendix @sec-app-alsampling, we repeat the analysis performed here on a different counterfactual consisting of plans sampled with a redistricting simulation algorithm; the results are broadly similar.

@@

This analysis is slightly more complicated than in New Jersey, since in order to understand the racial distribution of harm we need to estimate voting preferences by race.
As is custom in VRA analyses, we use an ecological inference model.
The model, detailed in Appendix @sec-app-alei, jointly estimates two turnout and vote choice by race and precinct across seven statewide elections using a setup following @king1997solution.
For example, we model precinct $j$'s observed turnout $n_{jv}$ in election $v$ and baseline turnout rates $\turn_{jr}$ by race $r$ as \begin{align*}
    n_{jv} \mid \turn_{jr}, \beta^{(t)}_v
        &\sim \Binlog\qty(
        \overbrace{\vap_j,}^{\substack{\text{voting-age}\\\text{population}}}
        \hspace{0.5em}
        \logit(\turn_j^\top
        \overbrace{\vb p_j}^{\mathclap{\substack{\text{precinct racial}\\\text{proportions}}}}) + 
        \underbrace{\beta^{(t)}_v}_{\mathllap{\text{election-specific turnout shift}}}) \\
    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
        &\sim \mathcal{N}\underbrace{\qty(\mu^{(t)}, \Sigma^{(t)})}_{\mathrlap{
        \text{where to cluster the precinct turnout rates}}},
\end{align*} where $\Binlog$ indicates the Binomial distribution with the probability parameter specified on the logit scale.
We model Democratic support rates $\supp_{jr}$ by precinct and race similarly.

@@

Figure @fig-boston-factors shows the key outputs from this model, and Appendix @sec-app-lfei explains the model in more detail.
As expected one dimension (the first) primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders, Elizabeth Warren, and current Boston mayor Michelle Wu on the left, and centrists like Joseph R. Biden and Wu's opponent, Annissa Essaibi-George, on the right.
The second dimension appears to capture racial differences in support---as the factor value maps in Figure @fig-boston-factors show, minority-majority and White areas of the city take opposite values on the second dimension.

Like all ecological inference models, ours relies on several unverifiable assumptions.
The latent factor component imposes additional assumptions about the structure of voting in Boston elections.
As Appendix @sec-app-lfei discusses, these assumptions are almost certainly not *strictly* met.
However, the model is a useful simplification of reality that captures the major electoral dynamics relevant to our question of interest.
For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure @fig-boston-gen-pred in Appendix @sec-app-lfei).
The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
We stress that the latent factor ecological inference model is not necessary for this kind of analysis; it is one out of several approaches which could provide estimated vote choices for individuals.
Local survey data with ideological preferences and racial information would be a sufficient alternative.

@@

```{=tex}
\vspace{1in}
\appendix
\renewcommand\thefigure{\thesection.\arabic{figure}}
\renewcommand\thetable{\thesection.\arabic{table}}
\setcounter{figure}{0}
\setcounter{table}{0}
```
# Two-party Election Model {#sec-app-elecmodel}

@@

# Correlation Between Partisan Gerrymandering Metrics Across Multiple States {#sec-app-corr}

@@

```{=tex}
\begin{table}[ht!]
\centering
\input{figures/corr_table.tex}
\vspace{0.1in}
\caption{Correlations between five gerrymandering metrics and partisan differential harm, for simulated districting plans from eight states. Declination is NA for Massachusetts since there are no Republican-leaning districts at baseline under any simulated plan.}
\label{tab:corr}
\end{table}
```
# Ecological Inference Model for Alabama {#sec-app-alei}

@@

# Additional Sampling Analysis for Alabama {#sec-app-alsampling}

@@

# Latent Factor Ecological Inference Model for Boston Primary Elections {#sec-app-lfei}

To estimate voter preferences by race and precinct in Boston, we use 2020 U.S.
Census demographic data and election returns from the March 2020 presidential primary and the September 2021 mayoral preliminary elections.
Our ecological inference model used to produce the estimates expands on the model in Appendix @sec-app-alei, jointly estimating both turnout and vote choice in the two elections.
However, it estimates voter preferences in a two-dimensional latent space and also estimates candidates' positions in this space.
